---
title: "P8105 HW3 | Jagjit Singh | ID: JS5958 "
output: github_document
date: "2022-10-11"
---

```{r setup}
library(p8105.datasets)
library(dplyr)
library(tidyverse)
library(ggridges)
library(grid)
library(gridExtra)
library(caret)
```



### Problem 1

#### Read in the data

```{r}
data("instacart")

instacart = 
  instacart %>% 
  as_tibble(instacart)
```

#### Answer questions about the data

This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are `r instacart %>% select(product_id) %>% distinct %>% count` products found in `r instacart %>% select(user_id, order_id) %>% distinct %>% count` orders from `r instacart %>% select(user_id) %>% distinct %>% count` distinct users.

Below is a table summarizing the number of items ordered from aisle. In total, there are 134 aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r}
instacart %>% 
  count(aisle) %>% 
  arrange(desc(n))
```

Next is a plot that shows the number of items ordered in each aisle. Here, aisles are ordered by ascending number of items.

```{r}
instacart %>% 
  count(aisle) %>% 
  filter(n > 10000) %>% 
  mutate(aisle = fct_reorder(aisle, n)) %>% 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

Our next table shows the three most popular items in aisles `baking ingredients`, `dog food care`, and `packaged vegetables fruits`, and includes the number of times each item is ordered in your table.

```{r}
instacart %>% 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>%
  group_by(aisle) %>% 
  count(product_name) %>% 
  mutate(rank = min_rank(desc(n))) %>% 
  filter(rank < 4) %>% 
  arrange(desc(n)) %>%
  knitr::kable()
```

Finally is a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week. This table has been formatted in an untidy manner for human readers. Pink Lady Apples are generally purchased slightly earlier in the day than Coffee Ice Cream, with the exception of day 5.

```{r}
instacart %>%
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_hour = mean(order_hour_of_day)) %>%
  spread(key = order_dow, value = mean_hour) %>%
  knitr::kable(digits = 2)
```


# PROBLEM 2
```{r}
accel_df = read_csv('/Users/jag/Downloads/accel_data.csv') %>%
  janitor::clean_names() 
```


```{r}
original_accel_df = 
  accel_df %>% dim()
```

```{r}

vars_accel = 
  accel_df %>% names()

dim_accel = 
  accel_df %>% dim()

modified_df <-
  accel_df %>%
  pivot_longer(
    activity_1:activity_1440,
    names_to = "minute",
    names_prefix = "activity_",
    values_to = "activity_count")

modified_df <- 
  modified_df %>%
  mutate(
    day = factor(day, levels = c("Sunday","Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")),
    minute = as.integer(minute),
    hour = as.integer(minute %/% 60), 
    week = as.integer(week), 
    weekday_vs_weekend = 
      case_when(
           day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday") ~ "Weekday", 
           day %in% c("Saturday", "Sunday") ~ "Weekend"),
    day_id = as.integer(day_id)) 


vars_modified = 
  modified_df %>% names()

dim_modified = 
  modified_df %>% dim()

```

We took the intial dataset and performs modifications (pivot, mutate) which would help us analyse the data and extract more meaning out of it. The modified data set had dimensions **`r dim_modified`** and whereas the original dataset has dimensions **`r dim_accel`**. 
The dataset is essentially the 5 weeks of readings from an accelerometer worn be 63 year-old male. The readings are the activity counts in a short period; one-minute intervals. From the modified data, which has a dimension of **`r vars_modified`** we can see that a total of 50400 minutes have been recorded which are the number of rows of the modified dataset due to our pivot. Our original dataset has dimension.
Also, the patient has been observed over a total of 35 days. We can verify it too as 24*35*60 = 50400. Also, there 1440 activites in the initial dataset which are the total number of minutes in a day. In terms of variables of the modified dataset. We have 


Now we aggregate across minutes to create a total activity variable for each day, and create a table showing these totals
```{r}
modified_df %>%
  rename(Day = day) %>% 
  group_by(Day) %>%
  summarize(Total_Activity_Count = sum(activity_count, na.rm = TRUE)) 
```
From the table above we can recognize a trend that total activity count is higher on Wednesday, Thursday, and Friday than compared to the other days. And we can see that the total activity is much lower on Saturday when compared to other days. This is plausible as Saturday is on a weekend, a holiday. 

Now we will make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week
```{r}
modified_df %>%
  group_by(day, minute) %>%
  ggplot(aes(x = minute, 
             y = activity_count, 
             color = day)) +
  geom_point() +
  labs(
    x = "Hour",
    y = "Net Count Per Hour")+ 
  scale_x_continuous(
    breaks = c(0, 240, 480, 720, 960, 1200, 1440),
    labels = c("0hr", "4hr", "8hr", "12hr", "16hr", "20hr", "24hr"))

```
From the above single panel scatter plot we can see that the highest activity count is around 8500. We can also see that the activity count is higher later at night around 9 pm on Friday, Saturday,Monday, Wednesday. During the first few days of the week, we can see that the activity count is higher around noon time. And we can see that the activity count is the lowest from around midnight till morning around 8 am. This gives an estimate of the patient's sleeping schedule .


# Problem 3

Loading the Dataset:
```{r}
library(p8105.datasets)
data("ny_noaa")
```

Data Cleaning:
```{r}
noaa_df <- 
  ny_noaa %>%
  janitor::clean_names() %>%
  mutate(tmax = as.double(tmax)/10, # converting back to Celsius
         tmin = as.double(tmin)/10, # converting back to Celsius
         prcp = as.double(prcp),
         snow = as.double(snow),
         snwd = as.double(snwd),
         id = as.factor(id))

noaa_df = drop_na(noaa_df)
```


Create separate variables for year, month, and day:
```{r}
modified_df <- 
  noaa_df%>% 
  separate(col = date, into = c("year","month", "day"))

```

Fetching summary statistics for each of the variables to get an estimate of the commonly observed values:
```{r}
hist(modified_df$snow)
```

From the histogram we can see that 0 is the most commonly observed value. 

# Subpart 2
Following is a two-panel plot showing the average max temperature in January and in July in each station across years. 
```{r}
modified_df$id = factor(modified_df$id)
modified_df$month = factor(modified_df$month)


sub_df_jan = subset(drop_na(modified_df), month == '01') %>%
  group_by(id) %>%
  summarize(mean_tmax = mean(tmax))


sub_df_july= subset(drop_na(modified_df), month == '07') %>%
  group_by(id) %>%
  summarize(mean_tmax = mean(tmax))

plot1 = ggplot(data = sub_df_jan,aes(x = id, y = mean_tmax)) + geom_bar(stat='identity', color = 'blue') +
  labs(title = 'January') +
    theme(
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

plot2 = ggplot(data = sub_df_july,aes(x = id, y = mean_tmax)) + geom_bar(stat='identity', color = 'pink') +
  labs(title = 'July') +
    theme(
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
grid.arrange(plot1,plot2, ncol=1)
```

The above two panel plot is a bar plot of the average max temperature across all the different station. Each bar represents a different station. We can see that can see that the January plot has a higher range difference than the July plot. This is plausible as weather tends have higher variation in winter months than summer. For January, we can see that 1 station has as mean tmax value greater than 7.5 and 1 station has mean tmax value less than -7.5. These can be considered outliers. There seems to be some structure in the July barplot, as all the mean tmax values are grate than 20 Degree Celsius. 



# Subpart 3 
Following is a two-panel plot showing: (i) tmax vs tmin for the full dataset; and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.

```{r}
plot1 = modified_df %>%
  select(tmax, tmin) %>% 
  pivot_longer(
    tmax:tmin,
    names_to = "observation", 
    values_to = "temp") %>% 
  ggplot(aes(x = temp, fill = observation)) +
  geom_density(alpha = .5) +
  viridis::scale_fill_viridis(discrete = TRUE)

subset_df_snow = subset(subset(modified_df, snow > 0),snow < 100)

subset_df_snow = subset_df_snow %>%
  select(id,year, snow) %>% 
  mutate(year = as.factor(year)) %>% 
pivot_longer(
    snow,
    names_to = "observation", 
    values_to = "value") 

plot2 = featurePlot(x = subset_df_snow %>%
dplyr::select(value),
y = subset_df_snow$year,
scales = list(x = list(relation = "free"),
y = list(relation = "free")),
plot = "density", pch = "|",
auto.key = list(columns = 2))

grid.arrange(plot1,plot2, nrow=1)
```
